---
title: "The htmlTable package"
author: "Max Gordon"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    css: custom.css
    keep_md: true
vignette: >
  %\VignetteIndexEntry{general}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

Basics
======

The **htmlTable** package is intended for generating tables using [HTML](http://en.wikipedia.org/wiki/HTML) formatting. This format is compatible with [Markdown](http://rmarkdown.rstudio.com/) when used for HTML-output. The most basic table can easily be created by just passing a `matrix` or a `data.frame` to the `htmlTable`-function:

```{r, results='asis'}
library(htmlTable)
# A simple output
output <- matrix(1:4,
                 ncol=2,
                 dimnames = list(list("Row 1", "Row 2"),
                                 list("Column 1", "Column 2")))
htmlTable(output)
```

Remember that you need to specify in your `knitr` Markdown have `results='asis'` at each chunk:

<pre>
```{r, results='asis', echo=FALSE}
cat("```{r, results='asis'}
   htmlTable(output)
```")
```
</pre>



Advanced
========

While it may be sufficient for basic tables a more advanced layout is often needed in medical publications with elements such as:

* row groups
* column spanners
* table spanners
* caption
* table footer
* zebra coloring (alias banding):
  + rows
  + columns

As many journals require that a MS Word-document is submitted it is furthermore also important that the table imports correctly to a word processor, i.e. that the table doesn't only look nice in a web browser but also in the final document. The `htmlTable`-function is written for all these purposes.

For demonstration purposes we will setup a basic matrix:

```{r}
mx <-
  matrix(ncol=6, nrow=8)
rownames(mx) <- paste(c("1st", "2nd",
                        "3rd",
                        paste0(4:8, "th")),
                      "row")
colnames(mx) <- paste(c("1st", "2nd",
                        "3rd", 
                        paste0(4:6, "th")),
                      "hdr")

for (nr in 1:nrow(mx)){
  for (nc in 1:ncol(mx)){
    mx[nr, nc] <-
      paste0(nr, ":", nc)
  }
}
```

Row groups
----------

The purpose of the row groups is to group variables that belong to the same group, e.g. a factored variable with more than two levels often benefit from grouping variables together.

```{r, results='asis'}
htmlTable(mx, 
          rgroup = paste("Group", LETTERS[1:3]),
          n.rgroup = c(2,4,nrow(mx) - 6))

```

We can easily mix row groups with regular variables by having an empty row group name `""`:

```{r, results='asis'}
htmlTable(mx, 
          rgroup = c(paste("Group", LETTERS[1:2]), ""),
          n.rgroup = c(2,4,nrow(mx) - 6))
```

When mixing row groups with variables without row groups we may want to omit the bold formatting of the row group label:

```{r, results='asis'}
htmlTable(mx, 
          css.rgroup = "",
          rgroup = c(paste("Group", LETTERS[1:2]), ""),
          n.rgroup = c(2,4,nrow(mx) - 6))
```

Column spanners
---------------

A column spanner spans 2 or more columns:

```{r, results='asis'}
htmlTable(mx,
          cgroup = c("Cgroup 1", "Cgroup 2"),
          n.cgroup = c(2,4))
```

It can sometimes be convenient to have column spanners in multiple levels:

```{r, results='asis'}
htmlTable(mx,
          cgroup = rbind(c("", "Column spanners", NA),
                         c("", "Cgroup 1", "Cgroup 2")),
          n.cgroup = rbind(c(1,2,NA),
                           c(2,2,2)))
```

Above example allows the column spanner to be a sum of the underlying cgroups (see n.cgroup), this is not required by the function:

```{r, results='asis'}
htmlTable(mx,
          cgroup = rbind(c("", "Column spanners", NA),
                         c("", "Cgroup 1", "Cgroup 2")),
          n.cgroup = rbind(c(1,5,NA),
                           c(2,1,3)))
```

Table spanners
--------------

A table spanner is similar to rgroup but has the primary purpose of combining 2 or more tables with the same columns into one:

```{r, results='asis'}
htmlTable(mx, 
          tspanner = paste("Spanner", LETTERS[1:3]),
          n.tspanner = c(2,4,nrow(mx) - 6))
```

Table caption
-------------

The table caption is simply the table description and can be either located above or below the table:

```{r, results='asis'}
htmlTable(mx[1:2,1:2], 
          caption="A table caption above")
```

```{r, results='asis'}
htmlTable(mx[1:2,1:2], 
          pos.caption = "bottom",
          caption="A table caption below")
```

A more interesting detail that the function allows for is table numbering, initialized by:

```{r}
options(table_counter = TRUE)
```

```{r, results='asis'}
htmlTable(mx[1:2,1:2], 
          caption="A table caption with a numbering")
```

As we often want to reference the table number in the text there are two associated functions:

```{r}
tblNoLast()
tblNoNext()
```

Table footer
------------

The footer usually contains specifics regarding variables and is always located at the foot of the table:

```{r, results='asis'}
htmlTable(mx[1:2,1:2], 
          tfoot="A table footer")
```


Zebra coloring (alias banded colors)
------------------------------------

Zebra coloring is also know as an alternating color pattern or row shading. It is most commonly applied to rows:

```{r, results='asis'}
htmlTable(mx, 
          col.rgroup = c("none", "#F7F7F7"))
```

The zebra coloring in `htmlTable` is unique in that it follows the rgroups. The zebra striping is centered around the rgroup although rows with no set rgroup, i.e. "" will have alternating colors event though they programatically are within the same group:

```{r, results='asis'}
htmlTable(mx, 
          col.rgroup = c("none", "#F7F7F7"),
          rgroup = c(paste("Group", LETTERS[1:2]), ""),
          n.rgroup = c(2,2,nrow(mx) - 4))
```

We can also color the columns:

```{r, results='asis'}
htmlTable(mx, 
          col.columns = c("none", "#F7F7F7"))
```

Or do both (note that the colors blend at the intersections):

```{r, results='asis'}
htmlTable(mx, 
          col.rgroup = c("none", "#F9FAF0"),
          col.columns = c("none", "#F1F0FA"))
```

Putting it all together
-----------------------

Now if we want to do everything in one table it may look like this:

```{r, results='asis'}
htmlTable(mx, 
          align="r",
          rgroup = paste("Group", LETTERS[1:3]),
          n.rgroup = c(2,4,nrow(mx) - 6),
          cgroup = rbind(c("", "Column spanners", NA),
                         c("", "Cgroup 1", "Cgroup 2&dagger;")),
          n.cgroup = rbind(c(1,2,NA),
                           c(2,2,2)),
          caption="A table with column spanners, row groups, and zebra striping",
          tfoot="&dagger; A table footer commment",
          cspan.rgroup = 2,
          col.columns = c(rep("none", 2),
                          rep("#F5FBFF", 4)),
          col.rgroup = c("none", "#F7F7F7"),
          css.cell = "padding-left: .5em; padding-right: .2em;")

```

